---
title: 'Chapter 7 Problems: Markov Chain Monte Carlo'
author: "Brandon Hoeft"
date: "April 8, 2018"
output:
  github_document:
    toc: TRUE
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = "",
                      fig.height = 6, fig.width = 6)
options(scipen=999) # turn of scientific notation.
```

## Exercise 7.1 Experiment with Metropolis Algorithm 

The script called `BernMetrop.R` from the program files for the book implements a Metrolopis Algorithm. Run the algorithm three times, each using different standard deviation of the **proposal distribution**. 

The **proposal distribution** enables Metropolis algorithm to take a random walk through the parameter space of $\theta$, effectively drawing samples from its posterior distribution . The SD of the **proposal distribution** we draw random samples from impacts how slowly or quickly the algorithm may explore the posterior distribution of $\theta$. 

Here we'll explore the impact of 3 different values of proposal distribution SD from a standardized normal distribution: 0.02, 0.2, 2.0.

### Metropolis Algorithm (MCMC) with Proposal SD of 0.02

First is using a proposal distribution SD of 0.02. We see the histogram of the posterior distribution followed by plots of the beginning and ending of the MCMC sampling chain. 

``` {r}
#source("/Users/bhoeft/Desktop/temp/DBDA Programs/DBDA2E-utilities.R"). 
# commented out line 4, 57. Change seed in line 55 to 123. comment out line 89 for openGraph
# so can generate the graph output in this RMarkdown script. 
source("/Users/bhoeft/Desktop/temp/DBDA Programs/BernMetrop_7_1_sd1.R")

```

### Metropolis Algorithm (MCMC) with Proposal SD of 0.2

``` {r}
# commented out line 4, 57. Change seed in line 55 to 123. comment out line 89 for openGraph
# so can generate the graph output in this RMarkdown script. 
source("/Users/bhoeft/Desktop/temp/DBDA Programs/BernMetrop_7_1_sd2.R")

```

### Metropolis Algorithm (MCMC) with Proposal SD of 2.0

``` {r}
# commented out line 4, 57. Change seed in line 55 to 123. comment out line 89 for openGraph
# so can generate the graph output in this RMarkdown script. 
source("/Users/bhoeft/Desktop/temp/DBDA Programs/BernMetrop_7_1_sd3.R")

```










